---
title: "Microarray data: Inspection"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "hide"
---

<style>
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
details p {font-family: monospace;}
</style>



## BDiB Session 04-March-2021  
### Overview  
In this workshop we will start by analysing a microarray dataset available on the [NCBI GEO](https://www.ncbi.nlm.nih.gov/geo/) repository.  
The dataset is identified by a ***data series*** (GSE) ID: GSE26868. The web page for this data series is [here](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE26868).  
The page contains quite a bit of information about the data, including e.g. the authors, a summary, a link to the PNAS paper where the data were first described.  
The page further provides access to the microarray data in different formats:  

* the raw (image) data: *GSE26868_RAW.tar*  
* processed data: normalised and log-transformed expression data: *"Series Matrix File"*  
* plus other data formats  

In this first script we will take a first look at the processed data matrix.  

```{r load packages, warning=FALSE, message=FALSE}
source("packages.R")
```


### Download data    
```{r download soft file}
dest.file <- "GSE26868_series_matrix.txt.gz"
if (!file.exists(dest.file)) {
  download.file(url=paste0(
    "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE26nnn",
    "/GSE26868/matrix/GSE26868_series_matrix.txt.gz"
  ),
  destfile = dest.file)
}
```


```{r file-name, results="asis"}
cat("Data file:<b>", dest.file, "</b>")
```


```{r open file, results="hide"}
gse <- GEOquery::getGEO(filename = dest.file, GSEMatrix=TRUE)
```

### Check what data class has been generated:  
```{r}
class(gse)
```

#### Probe IDs (print a few):
```{r}
head(Biobase::featureNames(gse))
```

### Sample info  
```{r}
Biobase::pData(gse)[,c("title", "source_name_ch1")] %>%
  knitr::kable(.) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped"))
```


### Plot data as a boxplot  
#### First obtain the data matrix:
```{r}
mx <- Biobase::exprs(gse)
# class(mx) # "matrix"
```


#### Print an extract:  
```{r}
mx[1:4, 1:4]
```
#### Generate plot:  
```{r message=FALSE, warning=FALSE}
op <- par()
par(mar=c(7,4,0,2)) # default is (5,4,4,2)
boxplot(mx, las=2)
par(op)
rm(op)
```
### Dimensionality reduction  
Each row in the data matrix represents one probe (>gene), and each column is one sample.

```{r results="asis"}
cat("Number of columns (samples): <b>", ncol(mx), "</b><br>\n",
    "Number of rows (probes): <b>", scales::comma(nrow(mx)), "</b>",
    sep="")
```

There are a number of different methods available for dimensionality reduction to compare similarities between the samples:  

#### Multidimensional scaling analysis (MDS)  
I recommend [this Youtube video](https://youtu.be/GEn-_dAyYME) for a nice explanation of MDS.

```{r fig.height=4, fig.width=6}
opar <- par(no.readonly = TRUE)
par(xpd = TRUE, mar = par()$mar + c(0, 0, 0, 5))
limma::plotMDS(mx, pch=1)
par(opar)
```

The plot above indicates three groups of samples. Let's check if they correspond to different cells types and/or treatments:  

```{r fig.height=4, fig.width=6}
cells <- Biobase::pData(gse)$characteristics_ch1 %>% 
  as.character() %>% 
  stringr::str_remove("^cell line: ") %>% 
  set_names(as.character(Biobase::pData(gse)$geo_accession))
labels.cells <- cells[colnames(mx)]

tnf <- grepl("tnf", Biobase::pData(gse)$description, ignore.case=T) %>% 
  as.numeric %>% add(1) %>% 
  set_names(as.character(Biobase::pData(gse)$geo_accession))
tnf <- tnf[colnames(mx)]
# unname(tnf) # 1 1 2 2 1 1 1 2 2 2
colrs <- c("brown", "blue")[tnf] # brown = ctrl; blue=TNF

opar <- par(no.readonly = TRUE) # saving current plotting parameters
par(xpd = TRUE, mar = par()$mar + c(0, 0, 0, 5))
limma::plotMDS(mx, labels=labels.cells, col=colrs)
legend(x="center", legend=c("Ctrl", "TNF"), pch=1,
       col=c("brown", "blue"), text.col = c("brown", "blue"))
par(opar) # restoring original plotting parameters
rm(cells, labels.cells, tnf, colrs, opar) # clear variables
```

The multidimensional scaling plot indicates that the samples cluster nicely by both cell type (greatest distance / dim 1) and treatment.

### Session info  
```{r SESSION INFO DATE, results="asis"}
cat("Date:", format(Sys.time(), "%a %d-%b-%Y %H:%M:%S"), "<br>\n")
```

```{r SESSION INFO PLATFORM}
si <- sessioninfo::session_info()
si.platform.df <- data.frame(
  setting = names(si$platform),
  value   = unlist(si$platform),
  stringsAsFactors = FALSE,
  row.names = NULL
)
DT::datatable(si.platform.df,
              options = list(columnDefs = list(list(className = 'dt-left', targets = 0:1)),
                             dom = 't'),
              rownames = FALSE,
              class="stripe")

```

```{r}
si$packages %>% 
  as.data.frame() %>% 
  dplyr::select(package, version=loadedversion, date, source) %>% 
  DT::datatable(options = list(columnDefs = list(list(className = 'dt-left', targets = 0:3))),
              rownames = FALSE,
              class="stripe")
  
```




