---
title: "Microarray data: Inspection"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "hide"
---


<style>
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
details p {font-family: monospace;}
</style>

## BDiB Session 04-March-2021  

```{r load packages, warning=FALSE, message=FALSE}
source("packages.R")
```


### Download data    
```{r download soft file}
dest.file <- "GSE26868_series_matrix.txt.gz"
if (!file.exists(dest.file)) {
  download.file(url=paste0(
    "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE26nnn",
    "/GSE26868/matrix/GSE26868_series_matrix.txt.gz"
  ),
  destfile = dest.file)
}
```


```{r file-name, results="asis"}
cat("Data file:<b>", dest.file, "</b>")
```


```{r open file, results="hide"}
gse <- GEOquery::getGEO(filename = dest.file, GSEMatrix=TRUE)
```

### Check what data class has been generated:  
```{r}
class(gse)
```

#### Probe IDs (print a few):
```{r}
head(Biobase::featureNames(gse))
```

### Sample info  
```{r}
Biobase::pData(gse)[,c("title", "source_name_ch1")] %>%
  knitr::kable(.) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped"))
```


### Plot data as a boxplot  
#### First obtain the data matrix:
```{r}
mx <- Biobase::exprs(gse)
# class(mx) # "matrix"
```


#### Print an extract:  
```{r}
mx[1:4, 1:4]
```
#### Generate plot:  
```{r message=FALSE, warning=FALSE}
op <- par()
par(mar=c(7,4,0,2)) # default is (5,4,4,2)
boxplot(mx, las=2)
par(op)
rm(op)
```
### Dimensionality reduction  
Each row in the data matrix represents one probe (>gene), and each column is one sample.

```{r results="asis"}
cat("Number of columns (samples): <b>", ncol(mx), "</b><br>\n",
    "Number of rows (probes): <b>", scales::comma(nrow(mx)), "</b>",
    sep="")
```

There are a number of different methods available for dimensionality reduction to compare similarities between the samples:  

#### Multidimensional scaling analysis (MDS)  
I recommend [this Youtube video](https://youtu.be/GEn-_dAyYME) for a nice explanation of MDS.

```{r fig.height=4, fig.width=6}
opar <- par(no.readonly = TRUE)
par(xpd = TRUE, mar = par()$mar + c(0, 0, 0, 5))
limma::plotMDS(mx, pch=1)
par(opar)
```

The plot above indicates three groups of samples. Let's check if they correspond to different cells types or treatments:  

```{r fig.height=4, fig.width=6}
cells <- Biobase::pData(gse)$characteristics_ch1 %>% 
  as.character() %>% 
  stringr::str_remove("^cell line: ") %>% 
  set_names(as.character(Biobase::pData(gse)$geo_accession))
labels.cells <- cells[colnames(mx)]

tnf <- grepl("tnf", Biobase::pData(gse)$description, ignore.case=T) %>% 
  as.numeric %>% add(1) %>% 
  set_names(as.character(Biobase::pData(gse)$geo_accession))
tnf <- tnf[colnames(mx)]
# unname(tnf) # 1 1 2 2 1 1 1 2 2 2
colrs <- c("brown", "blue")[tnf] # brown = ctrl; blue=TNF

opar <- par(no.readonly = TRUE)
par(xpd = TRUE, mar = par()$mar + c(0, 0, 0, 5))
limma::plotMDS(mx, labels=labels.cells, col=colrs)
legend(x="center", legend=c("Ctrl", "TNF"), pch=1,
       col=c("brown", "blue"), text.col = c("brown", "blue"))
par(opar)
rm(cells, labels.cells, tnf, colrs, opar)
```

#### Multidimensional scaling  
aka as *principal coordinates analysis*.  

```{r}
d <- dist(mx)
dim(d)
x <- d[, 1]
y <- d[, 2]
```





